Efficient forwards-time simulation with ancestral recombination graphs

### 2017-10-25 forwards_paper.tex
Added benchmarks and discussion.

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
A mutation occurs over a specific node at a given site, and results in a specific
derived state. Thus, all samples below the mutation node in the tree will inherit
this state (unless further mutations are encountered). Three mutations are shown
in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
mutation occurs at site zero (which is in the left-hand tree), which is a simple
mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
right hand tree) has two mutations: one occuring over node $3$ changing the state to
`C' and a back mutation over node $1$ changing the state to `G'.
\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
Perhaps if we added the state changes to the trees, and the states of the leaves
below the samples, we could get rid of all the example discussion here?}

\jk{Other stuff we can review here:}
\begin{enumerate}
\item Encoding a tree sequence topology like this is very efficient, much, much more compact than newick.
Can refer to example in msprime paper, or could do a large running example here. Say 500K samples of
human CHR1, something like that. Newick will be hundreds of TB.

\item Can also make the point that it's much more efficient than, say, VCF. We could drop some
mutations on the example above, and then report how big it would be if encoded with VCF and BCF.

\item The \msprime\ API is very efficient for working with these large datasets. Examples on
how long it takes to iterate over all the trees, compute pi, etc.

\item The point of all this is to emphasise that not only is it efficient to generate the
data as a tree sequence, it's also very efficient to store and process it this way too. By the
time we get to the end of this section we should have convinced the reader of this.

\end{enumerate}

>>>>

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
A mutation occurs over a specific node at a given site, and results in a specific
derived state. Thus, all samples below the mutation node in the tree will inherit
this state (unless further mutations are encountered). Three mutations are shown
in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
mutation occurs at site zero (which is in the left-hand tree), which is a simple
mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
right hand tree) has two mutations: one occurring over node $3$ changing the state to
`C' and a back mutation over node $1$ changing the state to `G'.
\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
Perhaps if we added the state changes to the trees, and the states of the leaves
below the samples, we could get rid of all the example discussion here?}

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million tree
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB. Working with either the Newick or VCF encoding
of this dataset would be exceedingly cumbersome, likely requiring several
days of CPU time simply to read the information info memory. \jk{Could
add concrete timing estimates based on subsets here, but maybe overdoing it?}

The tree sequence encoding is very compact but it is also very efficient to process.
Many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies. For example, the
pairwise nucleotide diversity $\pi$, is defined as the average number of
differences between sequences in the sample. Computing $\pi$ directly
from observed sequence data requires $O(n^2 m)$ time, since we must
compare all pairs of samples at every site. However, given the topology of
the trees at every site and the locations of mutations much better algorithms
are possible. Using the fact that we can count the number of samples below
a given node efficiently using previously described tree sequence
algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
becomes roughly $O(n \log n + m)$.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, and calculating $\pi$ over all 500K samples
in the example above required about 1.2 seconds.
\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
it too deeply. I guess we should also compare this with something else in order to
show that this time is pretty good. We could get a subset of the sites and push
the data into pylibseq maybe, or scikit-allel??}
